Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-05-30 manuscript/main.qmd
Lewis comments addressed until end of results

```{=html}
<!-- discarded titles:
Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene
A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->
```
Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate. Challenges arise particularly from the patchiness and uneven sampling of palaeoclimate data, and from the disparate nature of the proxy data. Geochemical data usually provide temperature estimates via transfer functions informed by experiments, and transfer functions based on the climatic requirements of modern taxa exist for some fossil plant assemblages. In contrast, most ecological and lithological data, e.g. coral reefs and evaporites, only convey information on broad climatic requirements. Historically, most large-scale proxy-based reconstructions have been based on either geochemical or ecological data, but few approaches exist for combining multiple proxy types into a single, quantitative reconstruction. Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into consideration the spatial distribution of the samples, leading to biased temperature reconstructions. Here, we present a Bayesian hierarchical model to integrate ecological data with established geochemical proxies into a unified quantitative framework, bridging gaps in the latitudinal coverage of proxy data. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data. Integrating multiple types of proxy data and adequate prior information has the potential to substantially reduce uncertainty in palaeoclimate reconstructions, allowing for unbiased temperature estimates from sparse data.

@@

Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018]. Recent advances in the synthesis of palaeoclimate data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, but a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures   [@judd2020; @Jones2022]. At the most basic level, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of other regions (e.g. a high proportion of samples from tropical areas). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions (e.g. @song2019). However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modeling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data. Statistical modeling in the Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. This model expands upon existing, spatially explicit palaeoclimate reconstructions  by allowing for the integration of (1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using downsampled, simulated temperature gradients.

We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data. We chose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018]. Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [@Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].

::: {#table-1}
![](../figures/Table1.png){width="16.25cm"}
Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies. For comparison, a gradient derived from a climate model ensemble is also shown.

@@

```
## Modern sea surface temperatures

## Geochemical data

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$. For our analyses, this dataset was restricted to the EECO (defined as 49.1 - 53.8 Ma) and samples from the continental shelf. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained. For a detailed description of each proxy see @Hollis2019.

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$N--32$^\circ$S), with minimum sea surface temperature tolerances (\~18$^\circ$C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6$^\circ$C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 $^\circ$N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34$^\circ$ N). Subsequently, we used statistically derived temperature limits (minimum = 21$^\circ$C, average = 27.6$^\circ$C, maximum = 29.5$^\circ$C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5$^\circ$C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. While factors beside sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred at polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa conform to similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator. For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates. From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6$^\circ$C (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*). However, a value of 22.5$^\circ$C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.

@@

**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (also known as "growth curve" or "Richard's curve") of the form:

@@

$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $${#eq-nu} where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, i.e. the latitude around which temperature falls most steeply with latitude, $B$ denotes the growth rate (Fig. 2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.

@@

**Priors.** In the Bayesian framework, priors need to be placed on the unknown parameters of a model. We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$: $$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $${#eq-sigma}
$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $${#eq-sigmaj}
We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.

In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and vaguely based on the modern climate system:

@@

**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient: $$M \sim N(42,10)$${#eq-M}

**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes. A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$${#eq-B}
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse [@Frakes1992]. We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types. Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, examplary distribution that captures both limited sample size and skewed geographic origins of samples. To assess how well the modeled gradients, reconstructed from limited sampling, fit the idealised gradients, we calculated an $R^2$ for Bayesian regression models [@Gelman2019]. For every iteration from the posterior, we intercepted the modeled and the idealised gradient in intervals of 1$^\circ$ latitude, and calculated the $R^2$ based on these values. We report the median, and $2.5^{th}$ and $97.5^{th}$ percentile of the resulting $R^2$ measures.

To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], down-sampled to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131). The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modeled gradient and the empirical temperature averages in 1$^\circ$ latitude bins. Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013]. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in. Every 10th iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation. The re-sampled, simulated gradients were modeled in one chain with 10,000 iterations for each of the 1,000 random samples. 5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs. For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Supplementary Figure S1).

@@

```{=html}
<!--
- Calculate lat gradient
- Calculate global T
[Fig. 4]
-->
```
Modeled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior. The median, $2.5^{th}$ and $97.5^{th}$ percentiles of temperatures where then taken from all temperature estimates obtained at the latitudes of interest.

The latitudinal gradient is calculated as the difference between the modeled temperature at the equator (0$^\circ$ latitude) and at the poles (90$^\circ$ absolute latitude). To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6$^\circ$ absolute latitude) being used instead of at the poles. Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig. 4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [@Evans2018].

Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.

Global average temperatures with 95% credible intervals were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1$^\circ$ latitudinal bins. The weights were set to the proportion of global surface area in each latitudinal bin, i.e. decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$${#eq-weights} where $\alpha_1$ is the upper, and $\alpha_2$ is the lower latitudinal boundary of bin $i$.

@@

## Model validation and modern gradient

@@

Present-day latitudinal temperature gradient. The present-day empirical latitudinal temperature gradient (mean sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise. Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modeled gradient.
:::

>>>>

```{=html}
<!-- discarded titles:
Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene
A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->
```
Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate. Particular challenges arise from the patchiness, uneven spatial distribution, and disparate nature of palaeoclimatic proxy records. Geochemical data typically provide temperature estimates via transfer functions derived from experiments. Similarly, transfer functions based on the climatic requirements of modern taxa exist for some fossil groups, such as pollen assemblages. In contrast, most ecological and lithological data (e.g. coral reefs and evaporites) only convey information on broad climatic requirements. Historically, most large-scale proxy-based reconstructions have used either geochemical or ecological data, but few studies have combined multiple proxy types into a single quantitative reconstruction. Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into account the spatial distribution of samples, leading to biased temperature reconstructions. Here, we present a Bayesian hierarchical model to integrate ecological data with established geochemical proxies into a unified quantitative framework, bridging gaps in the latitudinal coverage of proxy data. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures of the EECO. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data. We show that integrating multiple types of proxy data, and adequate prior information, has the potential to substantially reduce uncertainty in palaeoclimate reconstructions, allowing for unbiased temperature estimates from sparse data.

@@

Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018]. Recent advances in the synthesis of palaeoclimate data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, yet a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures   [@judd2020; @Jones2022]. Principally, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of other regions (e.g. a high proportion of samples from tropical areas). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions [e.g. @song2019]. However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data. Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using down-sampled, simulated temperature gradients.

We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data. We choose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018]. Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [Table 1; @Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].

::: {#table-1}
![](../figures/Table1.png){width="16.25cm"}
Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies. For comparison, a gradient derived from an atmosphere-ocean general circulation model ensemble is also shown.

@@

```
## Geochemical data

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$. For our analyses, this dataset was restricted to the EECO (defined as 53.8 -- 49.1 Ma) and samples from the continental shelf. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained. For a detailed description of each proxy see @Hollis2019.

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$ N -- 32$^\circ$ S), with minimum sea surface temperature tolerances (\~18$^\circ$C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6$^\circ$C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 $^\circ$ N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34$^\circ$ N). Subsequently, we used statistically derived temperature limits (minimum = 21$^\circ$C, average = 27.6$^\circ$C, maximum = 29.5$^\circ$C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5$^\circ$C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread and co-occurred across tropical and temperate latitudes in the early Eocene. Only *Avicennia*, however, occurred at polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa conform to similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator. For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates. From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6$^\circ$C (lower temperature limt of *Avicennia*) and 20.7$^\circ$C (lower temperature limit of *Rhizophora*). However, a value of 22.5$^\circ$C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.

@@

**Model structure.** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (also known as "growth curve" or "Richard's curve") of the form:

@@

$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $${#eq-nu} where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, i.e. the latitude around which temperature falls most steeply with latitude, $B$ denotes the growth rate, $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.

@@

**Priors.** In a Bayesian framework, priors need to be placed on the unknown parameters of a model. We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$: $$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $${#eq-sigma}
$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $${#eq-sigmaj}
We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.

In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and loosely based on the modern climate system:

@@

**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to 42, i.e. $M$ of the modern SST gradient, and a moderately wide standard deviation of 10: $$M \sim N(42,10)$${#eq-M}

**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to limit very abrupt SST changes across latitudes on a global scale. A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$${#eq-B}
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$. We informed the prior distributions on M and B based on a provisional model run with the modern SST data.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse, and extreme greenhouse [@Frakes1992]. We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types. Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, exemplary distribution that captures both limited sample size and skewed geographic origins of samples. To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019]. For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1$^\circ$ latitude and calculated the $R^2$ based on these values. We report the median, and 95% credible intervals (CI) of the resulting $R^2$ values. Here and in all other instances, the 95% CI refer to the interval between the 2.5% point and the 97.5% point of the samples or sampled posterior distribution.

To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], aggregated to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131). The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modelled gradient and the empirical temperature averages in 1$^\circ$ latitude bins. Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013]. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in. Every 10th iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation. The re-sampled, simulated gradients were modelled in one chain with 10,000 iterations for each of the 1,000 random samples. 5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs. For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Fig. S1).

@@

```{=html}
<!--
- Calculate lat gradient
- Calculate global T
[Fig. 4]
-->
```
modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior. The median and 95% CI of temperatures where then taken from all temperature estimates obtained at the latitudes of interest.

The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0$^\circ$ latitude) and at the poles (90$^\circ$ absolute latitude). To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6$^\circ$ absolute latitude) being used instead of at the poles. Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig. 4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [@Evans2018].

Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median (95% CI) from all pairs of iterations.

Global average temperatures with 95% credible intervals were calculated by taking the weighted mean of the median (95% CI) of temperature estimates in 1$^\circ$ latitudinal bins. The weights were set to the proportion of global surface area in each latitudinal bin, i.e. decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$${#eq-weights} where $\alpha_1$ is the upper, and $\alpha_2$ is the lower latitudinal boundary of bin $i$, i.e. we approximated the shape of the globe as a spheroid.

@@

@@

Present-day latitudinal temperature gradient. The present-day empirical latitudinal temperature gradient (median sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise. Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modelled gradient.
:::
